\section{Random Variation in Genetic Inheritance}
\label{sec:genetics}
Genetic variants are not randomly assigned, which means that simple correlations between the Ed PGI and observed outcomes cannot straightforwardly be interpreted as causal.
Genetic variation arises from inheritance: at conception, individuals inherit 23 chromosomes from each parent, obtaining half of their genetic variants from their mother and half from their father.
Because parents transmit genetic material to their children, those who inherit a higher Ed PGI typically also inherit familial environments shaped by parents who themselves possess a higher Ed PGI.
Consequently, correlations between an individual's Ed PGI and outcomes may reflect not only direct genetic effects but also indirect effects mediated through family environment, making these correlations confounded.

However, while genetic inheritance generates systematic associations between genes and environments, it also includes a random component.
Specifically, Mendelian inheritance ensures that, conditional upon parental genetics, there is random variation in the genetic variants a child inherits.
This provides an Instrumental Variable (IV) to isolate causal effects of Ed PGI on outcomes.
I exploit such Mendelian random variation to instrument for changes in the Ed PGI, thereby identifying causal relationships independent of confounding environmental factors.

\subsection{Mendelian Independent Inheritance}
I use variation away from parental values to estimate causal effects of having a higher versus lower Ed PGI.
This approach leverages the fundamental randomness in genetic inheritance, first described by Mendel's laws \citep{mendel1865versuche}.

Each parent has 0, 1, or 2 copies of each genetic variant --- corresponding to presence on each of their chromosomes.
When a father and mother's sex cells are formed, exactly one of the two copies is sorted into each reproductive cell, and is then inherited by their resulting child.
A parent with one copy therefore passes it on half the time, whereas a parent with 0 or 2 copies transmits their genetic variant with certainty.
At conception, the child receives one draw from the mother and one from the father, so the child's copy-count is simply the sum of two independent draws.

Write $x_{i,j}^\text{\textcolor{ForestGreen}{Child}}$ for the number of copies of variant $j$ that individual $i$ has (0, 1, or 2), and 
$x_{m(i),j}^\text{\textcolor{orange}{Mother}}, x_{f(i),j}^\text{\textcolor{blue}{Father}}$ for their respective parents $m(i), f(i)$.
If the mother has two copies of variant $j$ then they bequeath one copy to their child with certainty; if the father has one, then they bequeath one copy with probability $\frac 12$; if no copies, then they bequeath none with certainty.
%\[ \Egiven{x_{i,j}^\text{\textcolor{ForestGreen}{Child}}}{
%    x_{m(i),j}^\text{\textcolor{orange}{Mother}},
%        x_{f(i),j}^\text{\textcolor{blue}{Father}}}
%    = \Egiven{x_{i,j}^\text{\textcolor{ForestGreen}{Child}}}{
%        x_{f(i),j}^\text{\textcolor{blue}{Father}}}
%    + \begin{cases}
%        0, &\text{ if } x_{m(i),j}^\text{\textcolor{orange}{Mother}} = 0 \\
%        \frac 12, &\text{ if } x_{m(i),j}^\text{\textcolor{orange}{Mother}} = 1 \\
%        1, &\text{ if } x_{m(i),j}^\text{\textcolor{orange}{Mother}} = 2.
%    \end{cases} \]
The child then receives their other copy from their father, by the same process, giving the following representation in expectation,
\[ \Egiven{x_{i,j}^\text{\textcolor{ForestGreen}{Child}}}{
    x_{m(i),j}^\text{\textcolor{orange}{Mother}}, x_{f(i),j}^\text{\textcolor{blue}{Father}}}
    = \frac 12 \Big( x_{m(i),j}^\text{\textcolor{orange}{Mother}}
        + x_{f(i),j}^\text{\textcolor{blue}{Father}} \Big). \]

\begin{table}[!htbp]
    \centering
    \singlespacing
    \caption{Mendelian Independent Inheritance of One Genetic Variant.}
    \label{fig:mendelian-indep}
    %\input{sections/figures/mendelian-indep.tex}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Possible $x_{i,j}^\text{\textcolor{ForestGreen}{Child}}$, given parental values.}
        \begin{tabular}{l c | c c c }
            & & \multicolumn{3}{c}{$x_{f(i),j}^\text{\textcolor{blue}{Father}}$} \\
            & & 0         & 1      & 2       \\
            \midrule
            \multirow{3}{*}{$x_{m(i),j}^\text{\textcolor{orange}{Mother}}$} 
            & 0 & 0,0,0,0 & 0,1,0,1 & 1,1,1,1 \\
            & 1 & 0,0,1,1 & 0,1,1,2 & 1,1,2,2 \\
            & 2 & 1,1,1,1 & 1,2,1,2 & 2,2,2,2 \\
        \end{tabular}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Expected $x_{i,j}^\text{\textcolor{ForestGreen}{Child}}$, given parental values.}
        \begin{tabular}{l c | c c c }
            & & \multicolumn{3}{c}{$x_{f(i),j}^\text{\textcolor{blue}{Father}}$} \\
            & & 0         & 1      & 2       \\
            \midrule
            \multirow{3}{*}{$x_{m(i),j}^\text{\textcolor{orange}{Mother}}$} 
            & 0 & 0         & $0.5$ & 1       \\
            & 1 & $0.5$     & 1     & $1.5$   \\
            & 2 & 1         & $1.5$ & 2       \\
        \end{tabular}
    \end{subfigure}
    \justify
    \footnotesize
    \textbf{Note}:
    This figure shows the possible inherited values of one SNP, given parental values.
    The 4 values in the table (a) shows the 4 possible realisations that parents with variants $x_{m(i),j}^\text{\textcolor{orange}{Mother}}, x_{f(i),j}^\text{\textcolor{blue}{Father}}$.
    The values in table (b) show the expected value --- a simple mean of the 4 realisations in table (a).
\end{table}

After conception, the child has inherited genetic variants from both parents --- one of the 4 possible discrete realisations, not the expected value.
This result came about purely from random chance, conditional on at least one of parent having exactly one copy of the SNP, $x_{m(i),j}^\text{\textcolor{orange}{Mother}} = 1$ and/or $x_{f(i),j}^\text{\textcolor{blue}{Father}} = 1$.\footnote{
    Note both parents having 0 or 2 copies leaves $x_{i',j'}^\text{\textcolor{red}{Random}} = 0$, so that individual $i'$ inherited no random variation in variant $j'$.
}
\begin{equation}
    x_{i,j}^\text{\textcolor{red}{Random}}
    \coloneqq x_{i,j}^\text{\textcolor{ForestGreen}{Child}}
    - \Egiven{x_{i,j}^\text{\textcolor{ForestGreen}{Child}}}{
        x_{m(i),j}^\text{\textcolor{orange}{Mother}},
            x_{f(i),j}^\text{\textcolor{blue}{Father}}}
    \;\; \longrightarrow \;\; \text{Independently assigned.}
\end{equation}
$x_{i,j}^\text{\textcolor{red}{Random}}$ is the random variation in presence of SNP at position $j$.
From the possible values of the parents' variants, $x_{i,j}^\text{\textcolor{red}{Random}}$ equals zero if both parents have 0 or 2 copies, equals $\frac 12$ if the child actually inherited 1 copy when both parents had one copy each ($- \frac 12$ if they did not inherit), and 1 if they inherited 2 copies and the parents had one each ($-1$ if they inherited 0).

\subsection{Mendelian Independent Inheritance in the Ed PGI}
\label{sec:mendel-pgi}
Moving from a single variant to the PGI simply means summing across the same random-segregation for all the variants.
The Ed PGI is constructed by summing the 0, 1, or 2 copy indicators across the $J \approx 8,000$ different SNPs, each multiplied by the externally estimated weights $w_{j'}$ that capture how strongly each SNP predicts education years.
Because every variant copy enters the child's genome through an independent random draw (conditional on parental values), the child's Ed PGI is the weighted sum of thousands of such draws.

This means the Ed PGI can be decomposed into the component that was inherited as a result of a random draws from parent(s) who had 1 copy of the variant, and the certainty component from draws where parent(s) had 0 or 2 copies.
\[ Z_i^\text{\textcolor{ForestGreen}{Child}}
    = Z_i^\text{\textcolor{red}{Random}}
        + \Egiven{Z_i^\text{\textcolor{ForestGreen}{Child}}}{
            Z_{m(i)}^\text{\textcolor{orange}{Mother}},
                Z_{f(i)}^\text{\textcolor{blue}{Father}}}, \]
where $Z_i^\text{\textcolor{red}{Random}} = \sum_{j = 1}^J w_j x_{i,j}^\text{\textcolor{red}{Random}}$ is the random component of the Ed PGI for individual $i$, and similarly for their parents' Ed PGI values, $Z_{m(i)}^\text{\textcolor{orange}{Mother}}$ and $Z_{f(i)}^\text{\textcolor{blue}{Father}}$.
$Z_i^\text{\textcolor{red}{Random}}$ is the random component because it was the result of Mendelian independent inheritance, and its only determining factor was having at least one parent with exactly one copy for relevant genetic variants --- and nothing else.

Ideally, the random genetic component $Z_i^\text{\textcolor{red}{Random}}$ is constructed from observed parental Ed PGI values, $Z_{m(i)}^\text{\textcolor{orange}{Mother}}, Z_{f(i)}^\text{\textcolor{blue}{Father}}$.
However, parental PGI values are not observed for a large enough sample in the UKB.
Thus, I follow \cite{young2022mendelian} and leverage siblings' genetic data to impute parental PGI values.
This imputation recovers about 75\% of the parental genetic information, yielding imputed parental PGIs that are extremely highly correlated with the true parental values --- see \autoref{sec:pgi-impute}.

Formally, denote the Ed PGI value for individual $i$'s sibling $s(i)$ as $Z_{s(i)}^\text{\textcolor{purple}{Sibling}}$.
The observed random component, using sibling PGI values (instead of parents') is defined as follows,
\[ \tilde Z_i^\text{\textcolor{red}{Random}}
    = Z_i^\text{\textcolor{ForestGreen}{Child}}
    - \frac 12 \left( Z_i^\text{\textcolor{ForestGreen}{Child}} 
        + Z_{s(i)}^\text{\textcolor{purple}{Sibling}} \right)
    = \frac 12 \left( Z_i^\text{\textcolor{ForestGreen}{Child}} 
        - Z_{s(i)}^\text{\textcolor{purple}{Sibling}} \right). \]

        The imputed random component is equivalent to using the variation between an individual and their observed sibling.
This random component, therefore, is using the same within-family variation that early genetic studies did by examining within family differences.
An individual's Ed PGI is drawn from the same distribution as their sibling, because they had the same parents; in absence of observing both parents directly, this approach uses the sibling to infer this distribution.

This approach is equivalent, in expectation, to using observed parents because of the simple fact that children and parents are two independent draws from the parental genetic distribution.
Thus, in expectation, they both proxy for the parental PGI values,
\[ \E{\frac 12 \left( Z_i^\text{\textcolor{ForestGreen}{Child}} 
        + Z_{s(i)}^\text{\textcolor{purple}{Sibling}} \right)}
    = \frac12 \left(
        Z_{m(i)}^\text{\textcolor{orange}{Mother}}
            + Z_{f(i)}^\text{\textcolor{blue}{Father}} \right)
    \Rightarrow
    \E{ \tilde Z_i^\text{\textcolor{red}{Random}} } = 
        Z_i^\text{\textcolor{red}{Random}}. \]

By construction, $\tilde Z_i^\text{\textcolor{red}{Random}} $ captures Mendelian independent variation, conditional on the sibling's observed genetic values.
However, because siblings collectively reveal approximately 75\% of the parental genetic variation, the resulting imputed random component $\tilde Z_i^\text{\textcolor{red}{Random}}$ contains measurement error compared to the ideal (but unobserved) $Z_i^\text{\textcolor{red}{Random}}$.

The sibling-imputed parental mean PGI can be viewed as a proxy for the true parental mean PGI,
\[ \frac 12 \left( Z_i^\text{\textcolor{ForestGreen}{Child}} 
        + Z_{s(i)}^\text{\textcolor{purple}{Sibling}} \right)
    = \frac12 \left(
        Z_{m(i)}^\text{\textcolor{orange}{Mother}}
            + Z_{f(i)}^\text{\textcolor{blue}{Father}} \right) + \nu_i, \]
where $\nu_i$ denotes the measurement error arising from the incomplete observation of parental genotypes.
This error, $\nu_i$, comes from a sibling random inheriting more or fewer genes from one parent than the other, and is an independent draw from the primary individual by construction.
Notably, however, this error reduced precision of using the imputed proxy measure, compared true parental values

Consequently, the derived random genetic component $\tilde Z_i^\text{\textcolor{red}{Random}}$ maintains the same core identification properties as the ideal $Z_i^\text{\textcolor{red}{Random}}$ --- notably, Mendelian independent inheritance --- though it inherently includes measurement error due to sibling-based imputation. 
Its precision therefore depends on how genetically distinct each sibling pair is: sibling pairs who differ widely provide clearer signals, whereas genetically similar pairs introduce greater measurement noise.
Using $\tilde Z_i^\text{\textcolor{red}{Random}}$ as an IV yields consistent, but less precise estimates compared to the ideal scenario of directly observed parental genotypes.
For estimation, an Ordinary Least Squares (OLS) regression will implicitly weight observations by variance \citep{angrist1998estimating}, meaning sibling pairs with larger genetic differences disproportionately influence the estimated effects.
Thus, while the sibling-based imputation approach preserves unbiasedness in estimating genetic effects, it comes at the cost of increased standard errors and potentially reduced generalisability if sibling genetic divergence is not representative for the entire population.

\subsection{Measurement Error in the Ed PGI}
Previous studies have treated observed Ed PGI values as if they were perfect genetic measures, not accounting for the fact that $Z^\text{\textcolor{ForestGreen}{Child}}_i = \sum_{j'=1}^J w_{j'} z_{i,j'}$ is calculated from many statistical estimates of $w_{j'}$, which are themselves subject to measurement error.
Although the \cite{okbay2022polygenic} weights are estimates from an exceptionally large sample of over 3 million individuals, they remain subject to finite sample sampling variation and estimation error; the sample size is large, but the number of estimated parameters is similarly large ($w_{j'}$ for around 8,000 different values), making estimation error a sufficient concern.

Correspondingly, any statistical analysis using observed values of $Z^\text{\textcolor{ForestGreen}{Child}}_i$, or $\tilde Z_i^\text{\textcolor{red}{Random}}$, will suffer from attenuation bias and overconfident standard errors, which do not account for the statistical uncertainty in the GWAS weights $w_{j'}$.
This even holds true in an IV specification, because $Z^\text{\textcolor{ForestGreen}{Child}}_i$ and $\tilde Z_i^\text{\textcolor{red}{Random}}$ are measured with the same measurement error, because they are both functions of the GWAS weights $w_{j'}$.
In this case, the IV and endogenous variable have correlated measurement error, leading to attenuation bias in IV estimates.

To avoid this form of bias, I calculate $\tilde Z_i^\text{\textcolor{red}{Random}}$ using a set of GWAS weights estimated in a different population to that for $Z_i^\text{\textcolor{ForestGreen}{Child}}$.
Specifically, I take a secondary set of weights also provided in the supplementary data for \cite{okbay2022polygenic}, that were estimated exclusively from the UK Biobank subsample (and not the 23\&Me sample).\footnote{
    Note that this analysis also uses the UKB data, so that this secondary Ed PGI is calculated on a sample that overlaps the analysis sample.
    This will lead to over-fitting, although is ameliorated by the fact that the analysis sample is the sibling sample, about 10\% of the entire UKB, limiting the overlap to the extent possible.
}
This means uses an imputed random component which has measurement error coming from a different population than that for the Ed PGI, thus purging the attenuation bias in IV estimates.

This approach was first proposed in the genetics literature, to purge bias from PGI correlations \citep{van2023overcoming} and is applied here for the first time for causal estimates.
Using a PGI with one set of GWAS weights as an IV for another PGI constructed with different weights raises a key question: which set of weights should define the random (instrumental) component, and which should be used for the endogenous Ed PGI?
The method of Obviously Related Instrumental Variables (ORIV) removes this subjective choice by estimating the IV specification twice, once with each set of weights serving as the instrument for the other set \citep{gillen2019experimenting}.
ORIV then combines the estimates optimally and clustering the standard errors to account for the two-step estimation (i.e., clustering on individuals).
This yields unbiased estimates while efficiently accounting for estimation uncertainty from both sets of GWAS weights --- see \citealt{van2023overcoming} for full details.
% Appendix \ref{appendix:oriv} provides a detailed exposition of this econometric approach and its implementation in the context of PGIs.

\subsection{Causal Identification Strategy}
Formally, let $D_i(z')$ denote the potential education level and $Y_i(z', D_i(z'))$ labour market outcome for individual $i$, under the hypothetical alternative realisation that $i$ had Ed PGI value $Z_i^\text{\textcolor{ForestGreen}{Child}} = z'$.
Observed education and wages/income data are a sum of their respective potential outcomes.
\[ D_i = \sum_{z'} \indicator{
    Z_i^\text{\textcolor{ForestGreen}{Child}} = z'} D_i(z'), \;\;\;\;
    Y_i = \sum_{z'} \indicator{
    Z_i^\text{\textcolor{ForestGreen}{Child}} = z'} Y_i(z', D_i(z')).
\]
Note that $Y_i(z', D_i(z'))$ is a function of both hypothetical Ed PGI value, and resulting education $D_i(z')$; \autoref{sec:direct} returns to this distinction.

\noindent
\textbf{Target Parameters.}
This study aims to estimate the average causal effect of having a higher Ed PGI, which is summarised by the following average partial effects:
\[ \partialdiff{z'} \E{D_i(z')}, \;\;\;\; 
    \partialdiff{z'} \E{Y_i(z', D_i(z'))}. \]
The first average partial effect represents the average effect of a one standard deviation change in the Ed PGI on a measure of education, and the second on labour market wages or income.

The Ed PGI, however, is neither randomly assigned nor directly manipulatable, so that we may not observe $Y_i(z', D_i(z'))$ for different values of $z'$ for the same individual, and causal effects are not identifiable without further assumptions (i.e., the fundamental problem of causal inference).
Thus, I rely on random variation coming from the random component of the Ed PGI to infer causal effects.

\noindent
\textbf{Identifying Assumptions.}
I use the random component of the Ed PGI, $\tilde Z_i^\text{\textcolor{red}{Random}}$, as an instrument for the Ed PGI, $Z_i^\text{\textcolor{ForestGreen}{Child}}$.
This requires a number of assumptions to recover valid average causal effects.
\begin{enumerate}[label=(\arabic*)]
    \item Instrument Relevance and Independence: The random component of the Ed PGI strongly predicts the individual's Ed PGI, and is unrelated to potential outcomes.
    \item Instrument Monotonicity: 
    A higher value of the random component (weakly) increases the Ed PGI, with probability 1.
    \item Exclusion Restriction:
    The random genetic component, $\tilde Z_i^\text{\textcolor{red}{Random}}$, affects the potential outcomes only via the Ed PGI, $Z_i^\text{\textcolor{ForestGreen}{Child}}$.
\end{enumerate}

The three identifying assumptions are justified by the biology of genetic inheritance and the design of the empirical strategy.
The relevance and independence assumption (1) holds because the random genetic component arises solely from Mendelian inheritance, ensuring both predictive strength (see \autoref{tab:genetic-firststage}) and independence from unobserved confounding conditional on parental genetics and observed covariates.
The monotonicity assumption (2) is satisfied because inheriting a higher random genetic component can only additively increase the Ed PGI, never reducing it.
Indeed, the random component mechanically changes the Ed PGI for anyone who receives a higher random component, so that everyone is a complier, too.
Finally, the exclusion restriction (3) is credible given that genetic variants used in the random component influence socioeconomic outcomes only through their aggregation into the Ed PGI, by construction.

\noindent
\textbf{Estimation Strategy.}
I estimate the genetic first-stage as follows, 
\begin{equation}
    \label{eqn:genetic-firststage}
    Z_i^\text{\textcolor{ForestGreen}{Child}}
        = \gamma + \theta \tilde Z_i^\text{\textcolor{red}{Random}}
            + \vec\zeta' \vec X_i + \eta_i,
\end{equation}
where $\theta$ is the first-stage average contribution of the random component (with measurement error), and $\vec X_i$ is a vector of controls including parental imputed PGI levels, gender, age, sibling count, in city indicator, and the other PGIs (i.e., all demographic variables in \autoref{fig:demographic-correlates}).
Since $\tilde Z_i^\text{\textcolor{red}{Random}}$ serves as a formula instrument, it must be re-centred with respect to the control variables to produce valid causal effects \citep{borusyak2023nonrandom}.
I do this by calculating
$\Egiven{Z_i^\text{\textcolor{ForestGreen}{Child}}}{
    Z_{m(i)}^\text{\textcolor{orange}{Mother}},
        Z_{f(i)}^\text{\textcolor{blue}{Father}}}$
with interactions for the imputed parental PGI values and demographic variables.

I estimate the second-stage with a regression specification as follows, where $\vec X_i$ is a vector of control variables described previously.
\begin{align}
    \label{eqn:parametric-secondstage}
    Y_i &= \alpha + \beta Z_i^\text{\textcolor{ForestGreen}{Child}}
        + \vec \varphi' \vec X_i + \varepsilon_i.
\end{align}
$\beta$ represents the average effect of the Ed PGI on the education variables, and on wages and annual income (measured in $\log$ values).
I use the Ed PGI random component, $\tilde Z_i^\text{\textcolor{red}{Random}}$, to instrument for the Ed PGI in the IV specifications, using the ORIV method (with Ed PGI calculated using two independent weights) where mentioned.
The IV first-stage is shown in \autoref{eqn:genetic-firststage}, with the same set of control variables as the second-stage.

\noindent
\textbf{Limitations.}
The OLS and IV estimates place greater weight on individuals with larger deviations in the random component, $\tilde Z_i^\text{\textcolor{red}{Random}}$, which might not represent effects for the entire population --- potentially limiting external validity.
Additionally, despite independent Mendelian inheritance, the random genetic component remains modestly correlated with other PGIs.
While I control for these other PGI values, these correlations further constrain generalisability beyond the analysis sample, by implying overlap with the Ed PGI and these other PGIs.
Finally, the linear specifications employed (OLS and ORIV) yield weighted averages of individual-level causal effects.
This linear specification might not fully reflect heterogeneity in causal effects across the PGI distribution, or provide weights for the average causal effects representative for the entire UKB population --- see \cite{blandhol2022tsls} for further details.